Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-07 ms.Rnw
more work

When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is effaced, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
With sufficiently dense sampling, it can be possible to reconstruct the detailed history of who infected whom, which can aid identification of the correlates of transmission and immunity and the properties of infections.
More typically, one can use relatively sparse samples to learn about these features via the estimation of a parsimonious model of the transmission process.
In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
This is known as \emph{phylodynamic} inference, the term being due to \citet{Grenfell2004}.

The most common approach to phylodynamic inference rests upon a mathematical linkage between the tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among sampled genomes and a model of the dynamics of the transmission system.
Various linkages are possible, but because it is maximally efficient (\ie loses the least information), it is desirable to be able to compute the likelihood function for models of interest.
This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
To date, however, computation of this likelihood function has been out of reach, except in several special cases.

@@

In particular, existing approaches to likelihood-based phylodynamic inference have been based on one of two mathematical idealizations.
The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed via a reverse-time argument.
This computation is exact when the population size is constant.
Extensions of this approach develop approximate likelihoods for the case when the population size varies slowly \citep{Drummond2005} or when the sample fraction is negligible \citep{Volz2009a}.
The second idealization is the linear birth-death process, for which exact expressions for the likelihood are available \citep{Stadler2010}.
Linearity in this context amounts to the assumption that distinct lineages do not interact:
it is the resulting self-similarity of genealogies that renders the likelihood analytically tractable.
Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg]{MacPherson2021}.
Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.

@@

\citep{Alizon2024}

>>>>

When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is effaced, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
With sufficiently dense sampling, it can be possible to reconstruct the detailed history of who infected whom, which can aid identification of the correlates of transmission and immunity and the properties of infections.
More typically, one can use relatively sparse samples to learn about these features via the estimation of a parsimonious model of the transmission process.
In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
This is known as \emph{phylodynamic} inference, the term being due to \citet{Grenfell2004}.
\citet{Alizon2024} gives a good review of the history of the subject.

The most common approach to phylodynamic inference rests upon a mathematical linkage between the tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among sampled genomes and a model of the dynamics of the transmission system.
Various linkages are possible, but because it is maximally efficient (\ie loses the least information), it is desirable to be able to compute the likelihood function for models of interest.
This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
In particular, if $M$ represents a dynamic transmission model, $S$ a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, and $E$ a model of sequence evolution along a tree,
then the likelihood of $S$ is a function
\begin{equation*}
  f(S|M,E) = \int{f(S|\Phi,E)\,f(\Phi|M)\,\dd{\Phi}},
\end{equation*}
where we somewhat loosely use the functions $f$ to refer to distinct probability densities, the nature of each of which is clear from its arguments and the integral is taken over all possible genealogies relating the sequences $S$.
In this expression, $f(S|\Phi,E)$ is the \citet{Felsenstein2004} likelihood.
The function $f(\Phi|M)$, which links the phylogeny $\Phi$ to the dynamic model $M$, may be termed the \emph{phylodynamic likelihood}.
In the Bayesian context, this same function is sometimes referred to as a \emph{tree prior} \citep{Moeller2018,Volz2018}.
The computation of this function has remained out of reach, except in several special cases.

@@

In particular, existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed via a reverse-time argument.
This computation is exact when the population size is constant.
Extensions of this approach develop approximate likelihoods for the case when the population size varies slowly \citep{Drummond2005} or when the sample fraction is negligible \citep{Volz2009a}.
The second idealization is the linear birth-death process, for which exact expressions for the likelihood are available \citep{Stadler2010}.
Linearity in this context amounts to the assumption that distinct lineages do not interact:
it is the resulting self-similarity of genealogies that renders the likelihood analytically tractable.
Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg][]{MacPherson2021}.
Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.

@@
